%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Short Sectioned Assignment
% LaTeX Template
% Version 1.0 (5/5/12)
%
% This template has been downloaded from:
% http://www.LaTeXTemplates.com
%
% Original author:
% Frits Wenneker (http://www.howtotex.com)
%
% License:
% CC BY-NC-SA 3.0 (http://creativecommons.org/licenses/by-nc-sa/3.0/)
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%----------------------------------------------------------------------------------------
%	PACKAGES AND OTHER DOCUMENT CONFIGURATIONS
%----------------------------------------------------------------------------------------

\documentclass[paper=a4, fontsize=11pt]{scrartcl} % A4 paper and 11pt font size

\usepackage[T1]{fontenc} % Use 8-bit encoding that has 256 glyphs
\usepackage{fourier} % Use the Adobe Utopia font for the document - comment this line to return to the LaTeX default
\usepackage[english]{babel} % English language/hyphenation
\usepackage{amsmath,amsfonts,amsthm} % Math packages

\usepackage{lipsum} % Used for inserting dummy 'Lorem ipsum' text into the template

\usepackage{sectsty} % Allows customizing section commands
\allsectionsfont{\centering \normalfont\scshape} % Make all sections centered, the default font and small caps

\usepackage{fancyhdr} % Custom headers and footers
\pagestyle{fancyplain} % Makes all pages in the document conform to the custom headers and footers
\fancyhead{} % No page header - if you want one, create it in the same way as the footers below
\fancyfoot[L]{} % Empty left footer
\fancyfoot[C]{} % Empty center footer
\fancyfoot[R]{\thepage} % Page numbering for right footer
\renewcommand{\headrulewidth}{0pt} % Remove header underlines
\renewcommand{\footrulewidth}{0pt} % Remove footer underlines
\setlength{\headheight}{13.6pt} % Customize the height of the header

%\numberwithin{equation}{section} % Number equations within sections (i.e. 1.1, 1.2, 2.1, 2.2 instead of 1, 2, 3, 4)
\numberwithin{figure}{section} % Number figures within sections (i.e. 1.1, 1.2, 2.1, 2.2 instead of 1, 2, 3, 4)
\numberwithin{table}{section} % Number tables within sections (i.e. 1.1, 1.2, 2.1, 2.2 instead of 1, 2, 3, 4)

\setlength\parindent{0pt} % Removes all indentation from paragraphs - comment this line for an assignment with lots of text

%----------------------------------------------------------------------------------------
%	TITLE SECTION
%----------------------------------------------------------------------------------------

\newcommand{\horrule}[1]{\rule{\linewidth}{#1}} % Create horizontal rule command with 1 argument of height

\title{	
\normalfont \normalsize 
\textsc{AST4320} \\ [25pt] % Your university, school and/or department name(s)
\horrule{0.5pt} \\[0.4cm] % Thin top horizontal rule
\huge Assignment 3 \\ % The assignment title
\horrule{2pt} \\[0.5cm] % Thick bottom horizontal rule
}

\author{Ina Kullmann} % Your name

\date{\normalsize\today} % Today's date or a custom date

\begin{document}

\maketitle % Print the title

%----------------------------------------------------------------------------------------
%	PROBLEM 1
%----------------------------------------------------------------------------------------

\section{The isothermal density profile}

The second order differential equation for the density of the dark matter, $\rho(r)$ is given by:

\begin{equation}
-\frac{k_bT}{m_{DM}r^2} \frac{d}{dr}r^2 \frac{d}{dr} \ln \rho(r) = 4\pi G\rho(r)
\label{eq:DE}
\end{equation}

\textbf{Exercise:}  Show that $\rho(r) = \frac{A}{r^2}$, $A=\frac{k_bT}{2\pi Gm_{DM}}$ (the isothermal density profile) is a solution to the DE (that the DM settles into an isothermal density profile).\\

\textbf{Solution:} Rewriting the DE
\begin{align*}
-\frac{k_bT}{2\pi gm_{DM}} \frac{1}{r^2} \frac{d}{dr}r^2 \frac{d}{dr} \ln \rho(r) = -\frac{A}{r^2} \frac{d}{dr}r^2 \frac{d}{dr} \ln \rho(r) = 2\rho(r)
\end{align*}

Inserting the solution into the LHS of the DE:
\begin{align*}
-\frac{A}{r^2} \frac{d}{dr}r^2 \frac{d}{dr} \ln \Bigg(\frac{A}{r^2} \Bigg) &=  -\frac{A}{r^2} \frac{d}{dr}r^2 \frac{d}{dr} \Bigg( \ln A - \ln r^2 \Bigg) \\
&= \frac{A}{r^2} \frac{d}{dr}r^2 \frac{d}{dr} \Bigg( 2 \ln r \Bigg) \\
&= \frac{2A}{r^2} \frac{d}{dr}r^2 \cdot \frac{1}{r} \\
&= \frac{2A}{r^2} \frac{d}{dr}r \\
&= \frac{2A}{r^2} 
\end{align*}

Inserting the solution into the RHS of the DE:
\begin{align*}
2\rho(r) = \frac{2A}{r^2}
\end{align*}

\textbf{$\Rightarrow$ The LHS is equal to the RHS so that $\rho(r) = \frac{A}{r^2}$ is a solution to the DE.} \\

Gas (baryons) in hydrostatic equilibrium with gravity is described by:
\begin{equation}
\frac{dP}{dr} = -\frac{GM(<r) \rho_b(r)}{r^2}
\label{eq:hyd_eq}
\end{equation}
where the pressure $P$ is given by the baryons only, since the DM is pressureless. The right hand side is describing the gravitational pull from $M(<r)$, both the baryons and the DM, on the baryons $\rho_b$. \\

\textbf{Exercise:} Show that an isothermal gas (obeying equation \ref{eq:hyd_eq}) settles into a similar state (same density profile as the DM).\\

\textbf{Solution:} We want to show that the baryons settle into the same density profile as the DM, we want to show that
\begin{align*}
\rho_b(r) = \frac{B}{r^2}
\end{align*}
gives a solution to equation \ref{eq:hyd_eq}, where $B$ is a constant. If the gas is isothermal (it has the same temperature $T_0$ everywhere) then the pressure of the baryons are given by
\begin{align*}
P = nk_bT_0 = \frac{\rho_b}{m_p}k_bT_0 = \frac{k_bT_0B}{m_pr^2}
\end{align*}
giving the LHS:

\begin{align*}
\frac{d}{dr} \frac{k_bT_0B}{m_pr^2} = \frac{k_bT_0B}{m_p} \cdot -2r^{-3} = -\frac{2k_bT_0B}{m_pr^3}  
\end{align*}

The total mass enclosed in the sphere is given by the integral
\begin{align*}
M(<r) &= 4\pi \int_0^r x^2 \rho_{tot} (x) dx = 4\pi \int_0^r x^2 \bigg(\rho_b(x) + \rho_{DM}(x)\bigg) dx \\
&= 4\pi \int_0^r x^2 \Bigg(\frac{B}{x^2} + \frac{A}{x^2}\Bigg) dx = 4\pi (A+B)r
\end{align*}
so that the RHS becomes:
\begin{align*}
-\frac{GM(<r)\rho_b(r)}{r^2} = -\frac{G\cdot  4\pi (A+B)r \cdot \frac{B}{r^2}}{r^2} = -\frac{4\pi G(A+B)B}{r^3}
\end{align*}
setting the LHS equal to the RHS:
\begin{align*}
-\frac{2k_bT_0B}{m_pr^3} &= -\frac{4\pi G(A+B)B}{r^3} \\
\Rightarrow \frac{k_bT_0}{m_p} &= 2\pi G(A+B) \\
\Rightarrow  B &= \frac{k_bT_0}{2\pi G m_p} - A 
\end{align*}
\textbf{The baryons settle into the isothermal density profile if $B$ has the value given above, and that the DM also follows an isothermal density profile. }

%----------------------------------------------------------------------------------------
%	PROBLEM 2
%----------------------------------------------------------------------------------------

\section{The Ly$\alpha$ forest}

\textbf{Exercise:} Compute the Jeans length in the ionized IGM as a function of redshift. Assume $T_{IGM} = 10^4$K. What value of $k$ does this correspond to?

\textbf{Solution:} The Jeans mass and length is given by:
\begin{align*}
M_J &= \frac{4\pi}{3} \Bigg( \frac{\lambda_J}{2} \Bigg)^3 \rho_b = \frac{\pi^{5/2}}{6G^{3/2}\rho_b^{1/2}} \Bigg( \frac{k_bT}{\mu m_p} \Bigg)^{3/2} \\
\Bigg( \frac{\lambda_J}{2} \Bigg)^3 &= \frac{3}{4\pi}\frac{1}{\rho_b} \cdot \frac{\pi^{5/2}}{6G^{3/2}\rho_b^{1/2}} \Bigg( \frac{k_bT}{\mu m_p} \Bigg)^{3/2} \\
&= \frac{\pi^{3/2}}{8G^{3/2}\rho_b^{3/2}} \Bigg( \frac{k_bT}{\mu m_p} \Bigg)^{3/2} = \frac{1}{8} \Bigg(\frac{\pi}{G\rho_b}  \frac{k_bT}{\mu m_p} \Bigg)^{3/2}  \\
\lambda_J &= 2\cdot \Bigg( \frac{1}{8} \Bigg)^{1/3} \Bigg( \frac{\pi k_bT}{G\rho_b\mu m_p} \Bigg)^{1/2}
\end{align*}

from the lecture notes we know that
\begin{align*}
\bar{n}_H \approx 2 \cdot 10^{-7} (1 + z)^3 \mathrm{cm}^{-3}
\end{align*}
and that
\begin{align*}
\bar{n}_H  = f_H\frac{\rho_p}{m_p} = 0.76 \frac{\rho_b}{m_p} \Rightarrow \rho_b = \frac{m_p}{0.76}\bar{n}_H 
\end{align*}
inserting the numbers gives the Jeans length:

\begin{align*}
\lambda_J &= 2\cdot \Bigg( \frac{1}{8} \Bigg)^{1/3} \Bigg( \frac{\pi k_bT}{G \cdot \frac{m_p}{0.76}\bar{n}_H \cdot \mu m_p} \Bigg)^{1/2} \\
&= 2\cdot \Bigg( \frac{1}{8} \Bigg)^{1/3} \Bigg( \frac{0.76\pi k_bT}{G \bar{n}_H  \mu m_p^2} \Bigg)^{1/2} \\
&= 2\cdot \Bigg( \frac{1}{8} \Bigg)^{1/3} \cdot \frac{1}{m_p} \Bigg( \frac{0.76\pi k_bT }{G \cdot 2 \cdot 10^{-7} (1 + z)^3 \mathrm{cm}^{-3} \cdot  \mu} \Bigg)^{1/2} \\
&\approx 2\cdot \Bigg( \frac{1}{8} \Bigg)^{1/3} \cdot \frac{1}{m_p} \Bigg( \frac{0.76\pi k_bT }{G\cdot 2 \cdot 10^{-7}  \mathrm{cm}^{-3}  \mu} \Bigg)^{1/2} \cdot (1 + z)^{-3/2} \\
\end{align*}

This corresponds to a wavenumber $k$ given as:
\begin{align*}
k = \frac{2\pi}{\lambda_J}
\end{align*}

\textbf{Exercise:} In redshift-space, the Jeans length would correspond to a velocity width. What is this velocity width?\\

We want to calculate the width of the apsorption line (only) due to the expansion of the universe as the photons travel from the quazar to us. We assume that the IGM gas cloud has a final size that is roughly the Jeans length. We also assume that $\Omega_m = 1$ so that $H = H_0 \sqrt{\Omega_m(1+z)^3} = H_0 (1+z)^{3/2}$. Then the redshift space distortions are given by 
\begin{align*}
r = \lambda_J = \frac{\Delta \nu}{H(z)} = \frac{\Delta \nu}{H_0 (1+z)^{3/2}}
\end{align*}
giving the frequency width of the absorption line due the the expansion of the universe as:
\begin{align*}
\Delta \nu &= \lambda_J H_0 (1+z)^{3/2} = 2\cdot \Bigg( \frac{1}{8} \Bigg)^{1/3} \cdot \frac{1}{m_p} \Bigg( \frac{0.76\pi k_bT }{G\cdot 2 \cdot 10^{-7}  \mathrm{cm}^{-3}  \mu} \Bigg)^{1/2} \cdot (1 + z)^{-3/2} \cdot H_0 (1+z)^{3/2}  \\
&= 2\cdot \Bigg( \frac{1}{8} \Bigg)^{1/3} \cdot \frac{H_0}{m_p} \Bigg( \frac{0.76\pi k_bT }{G\cdot 2 \cdot 10^{-7}  \mathrm{cm}^{-3}  \mu} \Bigg)^{1/2} \\
\end{align*} 


\textbf{Exercise:} What does the finite width of the Ly$\alpha$ absorption cross-section do to absorption features in the Ly$\alpha$ forest? \\

The Ly$\alpha$ absorption line has a 'natural' with even if we ignore both the width due to the expansion of the universe and the width due to the thermal movement of the particles. This width is due to quantum mechanical effects of the absorption - we cannot know the wavelength/frequency of the line exact. This 'natural' line width is extremely small, in other words the cross section is extremely peaked (similar to a delta function). This narrow width can be exploited for instance when integrated over - we can replace the integral with a single value since the (delta) function only gives a non-zero value at the peak. \\

\textbf{Exercise:} Compare the thermal broadening scale to the Jeans length. \\

We want to calculate the width of the adsorption line (only) due to the thermal movement of the particles in the cloud. The standard deviation of the boltzmann distribution is given by
\begin{align*}
\sigma_v = \sqrt{\frac{3k_bT}{m}} 
\end{align*}
which would also correspond to the broadening of the width $\frac{\Delta \nu}{\nu}$ of the line due to the thermal motion of the gas particles





%----------------------------------------------------------------------------------------

\end{document}